Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-14 ind_files/03_BD_processes.tex
Significantly shorten and smoothen intro
Change filenames to be more sensible

\epigraph{\justifying Somewhere [ ... ] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding}

\section{Mathematical Background}

Here, I provide a brief, informal introduction to basic notions in stochastic processes that will be helpful for some technical portions of this thesis. Interested readers looking for a more comprehensive source can refer to standard mathematics texts such as \cite{oksendal_stochastic_1998} or \cite{karatzas_brownian_1998} for a more rigorous treatment of the mathematical foundations, or physics-style texts such as \cite{gardiner_stochastic_2009} or \cite{van_kampen_stochastic_1981} for useful tools and techniques to study real systems.

\subsection{Birth-death processes}
Mathematically, a birth-death process is a so-called `continuous-time Markov chain' in which only transitions between local states are allowed. In other words, a birth-death process is a stochastic process unfolding in continuous time such that
\begin{itemize}
	\item The process is `Markov', meaning that the future is statistically independent of the past given the present. In more mathematical terms, if the value of the stochastic process at time $t$ is given by $X_t$, $\mathbb{P}(\cdot | E)$ denotes probability conditioned on $E$, and $u < s \leq t$, then
	\begin{equation*}
	\mathbb{P}(X_t | X_s, X_u) = \mathbb{P}(X_t | X_s)
	\end{equation*}
	\item Direct transitions must be `local'. Mathematicians usually reserve the phrase `birth-death process' to processes that take values in the non-negative integers $\{0,1,2,3,4,\ldots\}$. In this case, only direct transitions from $n$ to $n \pm 1$ are allowed to occur. Biologically, this is saying that we observe the population on a fine enough timescale that the probability of two or more births/deaths occurring at the exact same time is very low and we can disallow it entirely in our models. The conditions for higher dimensional birth-death processes look similar.
\end{itemize}
Since these processes unfold in continuous time, they are characterized not by transition probabilities but by transition \emph{rates}, which can be thought of as the probability of transition `per unit time'. The quantity of interest is usually the probability of being in a particular state at a given point in time. The entire birth-death process can be described in terms of such a quantity, through a so-called `Master equation'. The master equation is a partial differential equation (PDE) for the probability of being in a given state at a given time, However, in all but the simplest cases, we can't actually solve this PDE, because it is simply too hard. The primary source of difficulty is non-linearity in the transition rates and the fact that transitions occur in discrete, discontinuous `jumps'. It is much easier to describe and analyze systems by using tools from stochastic calculus and partial differential equations, as we describe below.

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dB_s
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $B_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to B_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $B_t - B_s$ is independent of the past position $B_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $B_t - B_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$.\\
The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. Partition the interval $[0,T]$ into $n$ intervals of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:
\int\limits_{0}^{T} G(s,X_s)dB_s \coloneqq \lim_{n \to \infty} \sum\limits_{i=1}^{n}G(t_i,X_{t_i})(B_{t_{i+1}}-B_{t_i})  
That is to say, it is obtained by making successively finer partitions of the form $[t_i,t_{i+1}]$, and then computing the `area of the rectangle' formed with $B_{t_{i+1}}-B_{t_i}$ and $G(t_i,X_{t_i})$ as sides. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width corresponding to the random displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.\\
\\
Equation \eqref{ito_SDE_integral} is often represented in the compact form:
\label{ito_SDE_diff}
dX_t = F(t,X_t)dt + G(t,X_t)dB_t
. The physics literature also often uses the `Langevin form':
\label{ito_langevin}
\frac{dx}{dt} = F(t,x) + G(t,x)\eta(t)
where $\eta(t)$ is supposed to be `Gaussian white noise', defined indirectly such that $\int_0^{t}G(s,x)\eta(s)ds$ behaves identically to $\int_0^{t}G(s,X_s)dB_s$. However, it is important to remember that these are both purely formal expressions - Equation \eqref{ito_SDE_diff} is meaningless on its own and is really just shorthand for equation \eqref{ito_SDE_integral}, which is well-defined as explained above; Equation \eqref{ito_langevin} is even worse, because the Brownian motion is known to be non-differentiable, and as such, $\eta(t)$ cannot really exist - Both equations are thus to be interpreted as shorthand for equation \eqref{ito_SDE_integral}, which formally `makes sense'. SDEs are convenient because they satisfy several `nice' analytical properties. For example, using the fact that the Brownian motion has no expected change in value (\emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$), it can be shown\footnote{We can actually prove something stronger: We can show under rather mild regularity assumptions on $X_t$ and $G(t,x)$ that the stochastic integral is a continuous square-integrable martingale starting at the origin - This means that the map $t \to \int_0^{t}G(s,X_s)dB_s$ is continuous, starts at the origin, and always has an expectation value of $0$.} that the stochastic integral also has an expectation value of $0$ for all $t$, \emph{i.e}:
\mathbb{E}\left[\int\limits_{0}^{t} G(s,X_s)dB_s \bigg{|} X_0\right] = 0
Using this, and the fact that the future path of the Brownian motion itself is independent of its history, one can derive the following `notational algebra table' for manipulating products of formal expressions of the form \eqref{ito_SDE_diff}:
\\
\begin{center}
	\begin{tabularx}{0.4\textwidth}{ 
			| >{\centering\arraybackslash}X 
			| >{\centering\arraybackslash}X 
			| >{\centering\arraybackslash}X | }
		\hline
		& $\mathbf{dt}$ & $\mathbf{dB_t}$ \\
		\hline
		$\mathbf{dt}$ & $0$  &  $0$ \\ 
		\hline
		$\mathbf{dB_t}$ & $0$  & $dt$ \\
		\hline
	\end{tabularx}
\end{center}
which becomes very useful for formal manipulation. One important consequence is that we can no longer rely on the normal rules of calculus when dealing with stochastic integrals. In regular calculus, if we had a quantity $x(t)$ satisfying $\dot{x} = f(x,t) + g(x,t)$ for two `nice'\footnote{We just need the solution $x(t)$ to be a continuous function} real functions $f$ and $g$, then, given any function $h(x)$, we have the chain rule of differentiation, which says that
\frac{dh}{dt} = \frac{dh}{dx}\frac{dx}{dt} = h'(x)f(x) + h'(x)g(x)
\emph{i.e.}
	dh =  h'(x)f(x)dt + h'(x)g(x)dt
Naively, we may expect the same logic to still hold true for one-dimensional It\^o processes of the form \eqref{ito_SDE_diff} with $gdt$ simply being replaced by $GdB_t$ on the RHS. However, this does not work. The correct relation is instead given by 
\emph{It\^o's formula}\footnote{It\^o's formula also additionally requires $h \in C^2(\mathbb{R})$, meaning that $h$ is continuous and the first and second derivatives of $h$ exist and are also continuous}:
dh(X_t) = h'(X_t)F(X_t)dt + h'(X_t)G(X_t)dB_t  + \frac{h''(X_t)}{2}G^2(X_t)dt 
Note that there is now an extra $h''(X_t)G^2(X_t)/2$ term that does not exist in the deterministic setting. In some sense, this term is present because the random fluctuations of Brownian motion are `too erratic' and do not follow our deterministic intuitions. Using It\^o's formula and some simple algebra, one can then show that given any process $X_t$ taking values in $\mathbb{R}$ satisfying the SDE \eqref{ito_SDE_diff}, the associated probability density $P(x,t)$ of finding the process in a state $x \in \mathbb{R}$ satisfies the PDE
\label{ito_FPE}
\frac{\partial P}{\partial t}(x,t) = -\frac{\partial}{\partial x}\{F(t,x)P(x,t)\} + \frac{1}{2}\frac{\partial^2}{\partial x^2}\{(G(t,x))^2P(x,t)\}
I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature \citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics \citep{ewens_mathematical_2004} and pure mathematics \citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics \citep{ewens_mathematical_2004}, where it goes by the name of the `diffusion approximation', and has been heavily used by Kimura \citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation often lacks an explicit system size parameter (in physics parlance \citep{gardiner_stochastic_2009}, it is closer to a Kramers-Moyal expansion than a Van Kampen expansion) and is thus often somewhat ad-hoc.

\subsection{The intuition for the weak noise approximation in ecology}
If the parameter $K$ is sufficiently large, then the Fokker-Planck equation obtained via the system-size expansion can be further simplified to obtain a linear Fokker-Planck equation. This is accomplished by viewing the stochastic dynamics as fluctuating about a deterministic trajectory (obtained by letting $K \to \infty$) and only works if $K$ is large enough to be able to neglect all but the highest-order terms. This is usually an excellent approximation for populations in which the deterministic trajectory has already reached an attractor (stable fixed point, stable limit cycle, etc.). Since many deterministic eco-evolutionary models are expected to relax to such attractors, such an approximation is a useful first step in increasing the generality of existing models (which are usually studied only in the equilibrium regime) to incorporate the dynamics of finite populations. Importantly, this approximation \emph{only} works if we can discard all but highest-order terms of $K$: Including higher-order terms leads to equations that do not form Fokker-Planck equations and do not even describe probability densities. As such, this approximation is best suited to describe populations that are `medium sized' - small enough that they cannot be assumed to be infinitely large, yet large enough that stochasticity is rather weak and the deterministic limit is somewhat predictive - A situation that occurs frequently in ecology and evolution.

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. I will use this toy system as an illustration of the techniques that will be used for the actual problems we intend to tackle in the next sections. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.

\subsection{Description of the process and the Master Equation}

>>>>

\epigraph{\justifying Somewhere [...] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding}

Let us now consider a slightly more complicated scenario. Assume that our population is \emph{not} composed of identical organisms, but instead can contain up to $m$ different kinds of organisms - For example, individuals may come in one of $m$ colors, or a gene may have $m$ different alleles. The specific interpretation of the different variants is irrelevant to our formalism, and I, therefore, refer to each distinct variant of an organism simply as a `type'. Unlike many classic stochastic formulations in evolutionary theory~\citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate naturally over time.

\section{Description of the process and the Master Equation}
Given a population that can contain up to $m$ different (fixed) kinds of organisms, it can be entirely characterized by specifying the number of organisms of each type (Figure \ref{fig_nD_pop_description}A). Thus, the state of the population at a given time $t$ is an $m$-dimensional \emph{vector} of the form $\mathbf{n}(t) = [n_1(t),n_2(t),\ldots,n_m(t)]$, where $n_i(t)$ is the number of individuals of type $i$ in the population at time $t$.

Given a state $\mathbf{n}(t)$,  we also need to describe how this vector can change over time due to births and deaths (ecology). In this case, a birth or death could result in an individual belonging to one of $m$ different types. Thus, whereas before we had two functions $b(n)$ and $d(n)$ which take in a number as an input, we now require $2m$ functions that take in a vector as an input (Figure \ref{fig_nD_pop_description}B). In other words, for each type $i \in \{1,2,\ldots,m\}$, we must specify a birth rate $b_i(\mathbf{n})$ and a death rate $d_i(\mathbf{n})$. By `rates', we mean that if we know that \emph{either a birth or a death} occurs, then the probability that this event is the birth of an individual of type $i$ is given by
\begin{equation*}
\mathbb{P}\big[\textrm{ Birth of a type $i$ individual } \big{|} \textrm{ something happened }\big] = \frac{b_i(\mathbf{n})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{n})+d_j(\mathbf{n}))}
\end{equation*}
and the probability that the event is the death of an individual of type $i$ is
\begin{equation*}
\mathbb{P}\big[\textrm{ Death of a type $i$ individual } \big{|} \textrm{ something happened }\big] = \frac{d_i(\mathbf{n})}{\sum\limits_{j=1}^{m}(b_j(\mathbf{n})+d_j(\mathbf{n}))}
\end{equation*}

\myfig{0.9}{figures/3.1_BD_process_2D.png}{\textbf{Schematic description of a multi-dimensional birth-death process}. \textbf{(A)} Consider a population of birds in which individuals are either {\color{red}red} or {\color{blue}blue}. In this case, we have $m=2$, since there are two types of individuals in the population. \textbf{(B)} The state of the system can be described by a vector containing the number of individuals of each discrete type (in this case, the number of red and blue birds in the population). Births and deaths result in changes in the elements of the state vector.}{fig_nD_pop_description}

As before, we can describe the rate of change of $P(\mathbf{n},t)$, the probability of finding the population in a state $\mathbf{n}$ at time $t$, by measuring the inflow and outflow rates. Since the population changes in units of exactly one individual (by definition of a birth-death process; see section \ref{sec_math_background}), we know that these inflow and outflow rates must only involve populations that are a single individual away from our focal population. In other words, for a population $\mathbf{n} = [n_1,\ldots,n_{m}]$, the `inflow' is from all populations of the form $[n_1,\ldots,n_{i}-1,\dots,n_{m}]$ through a birth of a type $i$ individual, and from all populations of the form $[n_1,\ldots,n_{i}+1,\dots,n_{m}]$ through the death of a type $i$ individual. Thus, we have the inflow rate
\begin{equation}
\label{nD_rate_in}
\begin{split}
R_{\textrm{in}}(\mathbf{n},t) &= \sum\limits_{j=1}^{m}b_{j}([n_1,\ldots,n_{j}-1,\ldots,n_m])P([n_1,\ldots,n_{j}-1,\ldots,n_m],t) \\
& +\sum\limits_{j=1}^{m}d_{j}([n_1,\ldots,n_{j}+1,\ldots,n_m])P([n_1,\ldots,n_{j}+1,\ldots,n_m],t)
\end{split}
\end{equation}
Outflow is through births and deaths of individuals in the population $\mathbf{n}$ itself, and thus we have:
\begin{equation}
\label{nD_rate_out}
R_{\textrm{out}}(\mathbf{n},t) = \sum\limits_{j=1}^{m}b_{j}(\mathbf{n})P(\mathbf{n},t) + \sum\limits_{j=1}^{m}d_{j}(\mathbf{n})P(\mathbf{n},t)
\end{equation}
As before, we now define step operators, both for notational ease and in anticipation of the system size expansion. Note that now, we need $2m$ step operators. For each $i \in \{1,\ldots,m\}$, let us define two step operators $\mathcal{E}_{i}^{\pm}$ by their action on any function $f([n_1,\ldots,n_m],t)$ as:
\begin{equation}
\label{nD_step_operators}
\mathcal{E}_{i}^{\pm}f([n_1,\ldots,n_m],t) = f([n_1,\ldots,n_i \pm 1, \ldots n_m],t)
\end{equation}
In other words, $\mathcal{E}_{i}^{\pm}$ just changes the population through the addition or removal of one type $i$ individual. We can now the rate of change of $P(\mathbf{n},t)$ as
\begin{equation}
\frac{\partial P}{\partial t}(\mathbf{n},t) = R_{\textrm{in}}(\mathbf{n},t) - R_{\textrm{out}}(\mathbf{n},t)
\end{equation}
Substituting \eqref{nD_rate_in}, \eqref{nD_rate_out}, and \eqref{nD_step_operators}, we obtain:
\begin{equation}
\label{nD_M_eqn}
\frac{\partial P}{\partial t}(\mathbf{n},t) = \sum\limits_{j=1}^{m}\left[(\mathcal{E}_j^{-}-1)b_j(\mathbf{n})P(\mathbf{n},t) + (\mathcal{E}_j^{+}-1)d_j(\mathbf{n})P(\mathbf{n},t)\right]
\end{equation}
This is the master equation of our $m$-dimensional process.

\section{The system-size expansion}
As before, we now assume we can find a system size parameter $K > 0$ such that we can make the substitutions
\begin{align*}
\mathbf{x} &= \frac{\mathbf{n}}{K}\\
b^{(K)}_i(\mathbf{x}) &= \frac{1}{K}b_i(\mathbf{n})\\
d^{(K)}_i(\mathbf{x}) &= \frac{1}{K}d_i(\mathbf{n})
\end{align*}
and define new step operators $\Delta_{i}^{\pm}$ by their action on any real-valued function $f(\mathbf{x},t)$ as
\begin{equation}
\label{nD_step_operators_rescaled}
\Delta_{i}^{\pm}f([x_1,\ldots,x_m],t) = f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t)
\end{equation}
In terms of these new variables, \eqref{nD_M_eqn} becomes
\begin{equation}
\label{nd_M_eqn_rescaled}
\frac{\partial P}{\partial t}(\mathbf{x},t) = K\sum\limits_{j=1}^{m}\left[(\Delta_j^{-}-1)b^{(K)}_j(\mathbf{x})P(\mathbf{x},t) + (\Delta_j^{+}-1)d^{(K)}_j(\mathbf{x})P(\mathbf{x},t)\right]
\end{equation}
If $K$ is large, we can once again Taylor expand the action of the step operators as
\begin{equation*}
f([x_1,\ldots,x_i \pm \frac{1}{K}, \ldots x_m],t) = f(\mathbf{x},t) \pm \frac{1}{K}\frac{\partial f}{\partial x_i}(\mathbf{x},t) + \frac{1}{2K^2}\frac{\partial^2f}{\partial x_i^2}(\mathbf{x},t) + \mathcal{O}(K^{-3})
\end{equation*}
which, after substituting into \eqref{nd_M_eqn_rescaled}, yields the equation
\begin{equation}
\label{nD_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\mathbf{x},t) = \sum\limits_{j=1}^{m}\left[-\frac{\partial}{\partial x_j}\{A_j^{-}(\mathbf{x})P(\mathbf{x},t)\} + \frac{1}{2K}\frac{\partial^2}{\partial x_j^2}\{A_j^{+}(\mathbf{x})P(\mathbf{x},t)\}\right]}
\end{equation}
where
\begin{equation*}
A_{i}^{\pm}(\mathbf{x}) = b^{(K)}_i(\mathbf{x})\pm d^{(K)}_i(\mathbf{x})
\end{equation*}
Equation \eqref{nD_FPE} is an $m$-dimensional Fokker-Planck equation, and corresponds to the $m$-dimensional It\^o process
\begin{equation}
\label{nD_Ito_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where $\mathbf{A^-}(\mathbf{X}_t)$ is the $m$ dimensional `drift vector' with $i$\textsuperscript{th} element $ = A^{-}_{i}(\mathbf{X}_t)$. $\mathbf{D}(\mathbf{X}_t)$ is the $m \times m$ `diffusion matrix' with $ij$th element $\left(\mathbf{D}(\mathbf{X}_t)\right)_{ij} = \delta_{ij}\left(A^{+}_{i}A^{+}_{j}\right)^{\frac{1}{4}}$, where $\delta_{ij}$ is the Kronecker delta symbol, defined by
\begin{equation*}
\delta_{ij} = 
\begin{cases}
1 & i=j\\
0 & i\neq j
\end{cases}
\end{equation*}
Finally, $\mathbf{W}_t$ is the $m$-dimensional Wiener process and can be thought of as a vector of independent one-dimensional Wiener processes (which have been defined in \ref{intro_SDE}). This is the `mesoscopic' description of our process.

\section{Functional forms of the birth and death rates}

We assume that the birth and death rate functions have the functional form
\begin{equation}
\label{nD_functional_forms_for_replicator}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= x_ib^{\textrm{(ind)}}_{i}(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})
\end{aligned}
\end{equation}
where $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are non-negative functions that respectively describe the per-capita birth and death rate of type $i$ individuals, $\mu \geq 0$ is a constant describing the mutation rate in the population, and $Q_i(\mathbf{x})$ is a non-negative function that describes the additional birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$ that cannot be captured in the per-capita birth rate\footnote{When $x_i = 0$, \emph{i.e.} there are no type $i$ individuals in the population, individuals of type $i$ may still be born through mutations during births of the other types. This cannot be captured in $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ because the term $x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})$ vanishes when $x_i = 0$. Note that no analogous problem exists for the death rate, since the death rate of type $i$ individuals must be 0 when $x_i$ is 0 to ensure that we never have negative population densities.}. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates for each type. We define the \emph{Malthusian fitness} of the $i\textsuperscript{th}$ type as $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$, and the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type as $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + d^{\textrm{(ind)}}_{i}(\mathbf{x})$.  The quantity $w_i(\mathbf{x})$ describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$ discounting mutation. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that both $w_i$ and $\tau_i$ depend on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and not just on the density of the focal type. Thus, in general, both the fitness and the turnover rate in our model are frequency-dependent.

\section{Statistical measures for population-level quantities}\label{sec_stat_measures}
Given a state $\mathbf{x}(t)$ that describes our population at time $t$, we can compute the total (scaled) population size and the frequency of each type in the population as:
\begin{equation}
\label{nD_tot_pop_and_prop_inds_defn}
\begin{aligned}
N_{K}(t) &\coloneqq \sum\limits_{i=1}^{m}x_i(t) = \frac{1}{K}\sum\limits_{i=1}^{m}v_i(t)\\
p_i(t) &\coloneqq \frac{x_i(t)}{N_{K}(t)}
\end{aligned}
\end{equation}
These definitions are important because evolution is often measured as changes in type frequency (and \emph{not} necessarily changes in type density). Note that type frequency is a population level. Indeed, even though the causes of evolution are generally at the individual level, we are very often interested in quantities \emph{described} at a population level, such as the mean fitness or mean phenotype in the population~\citep{bourrat_evolution_2019}. Furthermore, the relevant quantities at the individual level, such as individual fitness or phenotype, are typically equal for all individuals of the same type (in some sense this is our basis for defining different types in the first place), and we, therefore, call them `type-level' quantities below. Let $f$ be any type-level quantity with (possibly time-dependent) value $f_i$ for the $i\textsuperscript{th}$ type. For example, if each type is a phenotype for a trait such as height, which can be assigned a numerical value, then setting $f_i = \textit{value of $i\textsuperscript{th}$ phenotype}$ gives us the mean trait value in the population. To facilitate the description of such quantities, we define the statistical mean value of any quantity $f$ in the population as
\begin{equation}
\label{nD_mean}
\overline{f}(t) \coloneqq \sum\limits_{i=1}^{m}f_ip_{i}
\end{equation}
the statistical covariance of two such quantities $f$ and $g$ as
\begin{equation}
\label{nD_cov}
\textrm{Cov}(f,g) \coloneqq \overline{fg} - \overline{f}\overline{g}
\end{equation}
and the statistical variance of a quantity $f$ as $\sigma^2_f \coloneqq \textrm{Cov}(f,f)$. It is important to recognize that these quantities are distinct from and independent of the \emph{probabilistic} expectation, variance, and covariance obtained by integrating over realizations in the underlying probability space. Indeed, for finite populations, the statistical mean, statistical variance, and statistical covariance are all themselves stochastic processes: For each instant of time, these population-level quantities are a random variable (\emph{i.e.} a \emph{function} and not just a number) depending on $\mathbf{p}$, the (random) vector of type frequencies in the population. For infinite populations, the statistical mean, variance, and covariance are entirely deterministic time-dependent quantities that simply describe how $f$ is distributed across the population. Failure to clearly make this distinction between statistical operations and probabilistic operations has led to much confusion with regard to the infinite population Price equation~\citep{van_veelen_use_2005}, which is entirely deterministic.

\section{Stochastic Trait Frequency Dynamics}

In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike~\citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation.  Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, we show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes and we have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. We will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in Chapter \ref{chap_unification}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'.

\section{The infinite population limit}\label{sec_nD_det_limit}
